---
title: 'Adam Zeilinger ZIB model (v.3) Month Fixed'
author: Daniel Turek
publish: true
---



## Updates (v.5)

- This analysis consists of Part III, using seasonal fixed effects.

- Added monitors to record samples for p\_occ and p\_obs.



## Updates (v.3)

- Changed the month covariate into a factor, that is, a fixed value 1, 2, 3, ..., 12 corresponding to each month, instead of a standardized continuous variable as before.

- Part I: Used this month factor as a fixed effect in the model, and removed the old month variables: month, quadratic month, and interactions between month and year

- Part II: Added a second model, *without* the monthly fixed effets, since none of them had any significant effect.

- Also included 90% credible intervals for the second model (Part II), since several of the `beta[i]` become significant at the 90% level.



## Updates (v.2)

- Using new data set with N = 958 observations (lists >= 4 species)

- Removed quadratic year^2 effect, and added year:month and year:month^2 interactions to occpancy process

- Transformed new siteID (raster cell numbers) into sequential indexing of random effects

- Changed the prior for sigma_alpha form uniform(0, 500) to uniform(0, 1000), to capture more of the long right-tail of the posterior.




## Preliminaries

Now using the seasonal effects data set.


```{r, message = FALSE}
require(nimble)
require(rjags)
require(coda)

## this is the new data set, with season as a factor: 1, 2, 3, 4
load('../data/zib_data_v5_season.Rdata')

source('definitions.R')
```





##Part III: Season as a fixed effect


Same as before, except season will be a fixed effect.

```{r define-dOcc-model-season}
code <- nimbleCode({
    mu_alpha ~ dnorm(0, 0.001)
    sigma_alpha ~ dunif(0, 1000)
    for(j in 1:nsite) { 
        alpha[j] ~ dnorm(mu_alpha, sd = sigma_alpha)  ## site random effect
    }
    for(i in 1:7) {
        beta[i] ~ dnorm(0, 0.001)
    }
    for(i in 1:4) {
        betaseason[i] ~ dnorm(0, 0.001)    ## new fixed effects for each season
    }
    for(i in 1:N) {
        logit(p_occ[i]) <- alpha[siteID[i]] + beta[4]*aet[i] + beta[5]*tmn[i] + beta[6]*tmx[i] + beta[7]*year[i] + betaseason[season[i]]
        logit(p_obs[i]) <- beta[1] + beta[2]*list_length[i] + beta[3]*year_list_length[i]
        y[i] ~ dOccupancy(p_occ[i], p_obs[i])
    }
})

constants <- list(N=N, nsite=nsite, aet=aet, tmn=tmn, tmx=tmx, year=year, season=season, list_length=list_length, year_list_length=year_list_length, siteID=siteID)

data <- list(y=y)

inits <- list(mu_alpha=0, sigma_alpha=1, alpha=rep(0,nsite), beta=rep(0,7), betaseason=rep(0,4))

modelInfo_season <- list(code=code, constants=constants, data=data, inits=inits, name='season')
```





## Part III: MCMC runs with seasonal fixed effects


```{r, eval = FALSE}
Rmodel <- nimbleModel(modelInfo_season$code,
                      modelInfo_season$constants,
                      modelInfo_season$data,
                      modelInfo_season$inits)

Cmodel <- compileNimble(Rmodel)

spec <- configureMCMC(Rmodel)

## configure this specification the same as log_shft_blk as before
spec$removeSamplers('beta[1:7]')
spec$addSampler('beta[1:3]', 'RW_block')
spec$addSampler('beta[4:7]', 'RW_block')
spec$removeSamplers('sigma_alpha')
spec$addSampler('sigma_alpha', 'RW_log_shift', list(shiftNodes='alpha'))

## additional additional monitors for p_occ and p_obs
spec$addMonitors(c('p_occ', 'p_obs'))

Rmcmc <- buildMCMC(spec)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

Now we'll run two chains of this MCMC algorithm, each with 500,000 MCMC iterations.

We'll manually remove the first 100,000 samples as burnin.

```{r, eval = FALSE}
niter <- 500000

set.seed(1)
Cmcmc$run(niter)
samples1 <- as.matrix(Cmcmc$mvSamples)[100001:niter, ]

set.seed(2)
Cmcmc$run(niter)
samples2 <- as.matrix(Cmcmc$mvSamples)[100001:niter, ]

mcmc1 <- coda::as.mcmc(samples1)
mcmc2 <- coda::as.mcmc(samples2)
mcmcs <- coda::mcmc.list(mcmc1, mcmc2)

save(samples1, samples2, mcmcs, file = '../cached/final_MCMC_fixedmonth.RData')
```

```{r, echo = FALSE}
load(file = '../cached/final_MCMC_fixedmonth.RData')
```


## Part III: Assessing convergence

Now we assess convergence using these two MCMC chains.

We'll use the mainstream Gelman & Rubin convergence diagnostic from the `coda` package.  We want to see all values near to one.

```{r gelman-diag}
coda::gelman.diag(mcmcs, autoburnin = FALSE)
```

Convergence looks excellent.

```{r final-mcmc-ess}
apply(samples1, 2, coda::effectiveSize)
```





## Part III: Posterior Inferences

### Part III: Posterior Mean

```{r}
cbind(apply(samples1, 2, mean),
      apply(samples2, 2, mean))
```

### Part III: Posterior Median

```{r}
cbind(apply(samples1, 2, median),
      apply(samples2, 2, median))
```

### Part III: 95% Credible Intervals

```{r}
cbind(apply(samples1, 2, function(x) quantile(x, 0.025)),
      apply(samples1, 2, function(x) quantile(x, 0.975)))
```




### Part II: Posterior Density Plots

```{r, eval = FALSE}
plot(mcmcs[[1]], ask = FALSE)
```

```{r posterior-plots-no-month, echo = FALSE}
for(i in 1:5)
    plot(coda::as.mcmc(samples1[, (3*i-2):min(3*i, dim(mcmcs[[1]])[2])]), ask = FALSE)
```








